Modelling state-switching functional data with hidden Markov models

### 2020-11-06 Report/Sections/models.tex
Fix Overleaf issues

In many applications where data are collected at high frequencies, intricate dependency structures may arise within the fine-scale process which cannot be adequately modelled with any of the HMM-based models mentioned thus far. For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour. To handle these additional fine-scale structures, we recommend replacing $Y_t^* = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant statistics that summarize this non-Markovian behaviour. To maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over the elements $Y_t^*$. Subject-area experts are often required to determine the optimal window size and the specific summary statistics used. For our case study, we use the discrete Fourier transform (DFT) of a moving forward window of width $h$ across $Y^*_t$. For each $t^* = 1,2,\ldots,T^*_t-h+1$:
%
\begin{align*}
    DFT\{Y^*_{t,t^*},\ldots, Y^*_{t,t^*+h-1}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right) \qquad 
\end{align*}
%
for $k = 0, 1, \ldots, h-1$. If $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise. Next, we calculate $\Z_{t,t^*} \equiv \{\Zone_{t,t^*},\Ztwo_{t,t^*}\}$ for each moving window, where
%
\begin{equation}
    \label{eqn:z}
    \Zone_{t,t^*} \equiv \frac{1}{h}\sum_{n=0}^{h-1}Y^*_{t,t^*+n}, \quad \Ztwo_{t,t^*} \equiv \sum_{k=1}^{\tilde{\omega}}\Big|\Big|DFT\{Y^*_{t,t^*},\ldots, Y^*_{t,t^*+h-1}\}(k)\Big|\Big|^2.
\end{equation}
%
In words, $\Zone_{t,t^*}$ is the average value over the window and $\Ztwo_{t,t^*}$ is the squared two-norm of the component of the window $\big\{Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\big\}$ that can be attributed to frequencies between one and $\tilde{\omega} \leq h-1$ periods per window (by the Plancherel theorem). More intuitively, $\Ztwo_{t,t^*}$ corresponds to the ``wiggliness" of the acceleration data for dive $t$ within the window starting at time stamp $t^*$. The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with the help of a subject-matter expert. Other DFT-based choices for window summary statistics are possible, including the dominant frequency and amplitude.

To avoid introducing auto-correlation into the transformed observation sequence, we recommend down-sampling $\Z_t$ such that $\Z_t = \big\{\Z_{t,1},\Z_{t,h+1},\Z_{t,2h+1},\ldots\big\}$ and $X^*_t = \big\{X^*_{t,1},X^*_{t,h+1},X^*_{t,2h+1},\ldots\big\}$. When down-sampling, we redefine the fine-scale probability transition matrices such that $\Gamma^{*(i)}_{jk} = \Pr(X^*_{t,t^*+h} = k | X^*_{t,t^*} = j, X_t = i)$ for all $t$ and $t^* = 1,h+1,2h+1,\ldots$. If the fine-scale model is a CarHMM, we also redefine the probability density function $f$ such that $f^{*(i,i^*)}(\cdot|\z_{t,t^*-100}; \theta^{*(i,i^*)})$ describes the density of $\Z_{t,t^*}$ given $X_t = i$, $X^*_{t,t^*} = i^*$, and $\Z_{t,t^*-100} = \z_{t,t^*-100}$.
This down-sampling both avoids the artificial residual correlation that overlapping windows would introduce and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting. By summarizing the data using a relevant transformation, the method overall also reduces correlation in the raw observations due to the complex dependence structure. Figure \ref{fig:models}d represents the process of transforming $Y^*_{t}$ into $\Z_{t}$ within an HMM, including down-sampling.

Once $\Z_t$ is calculated, it simply replaces the original fine-scale observations $Y^*_t$ in the fine-scale HMM structure. As such, the emission distributions and the likelihood remain identical to the original HMM when transforming $Y^*$ to $\Z$. In order to clearly differentiate models, we refer to an HMM with $\Z$ as observations as an HMM-DFT.

>>>>

In many applications where data are collected at high frequencies, intricate dependency structures may arise within the fine-scale process which cannot be adequately modelled with any of the HMM-based models mentioned thus far. For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour. To handle these additional fine-scale structures, we recommend replacing $Y_t^* = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant statistics that summarize this non-Markovian behaviour. To maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over the elements $Y_t^*$. Subject-area experts are often required to determine the optimal window size and the specific summary statistics used. For our case study, we use the discrete Fourier transform (DFT) of a moving forward window of width $h$ and stride $h$ across $Y^*_t$. Namely:
%
\begin{align*}
    DFT\{Y^*_{t,t^*},\ldots, Y^*_{t,t^*+h-1}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right)
\end{align*}
%
for $t^* = 1,h+1,2h+1,\ldots$ and for $k = 0, 1, \ldots, h-1$. If $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise. Next, we index each window with $\tilde{t}^*$ and calculate the transformed observations $\Z_{t,\tilde{t}^*} \equiv \{\Zone_{t,\tilde{t}^*},\Ztwo_{t,\tilde{t}^*}\}$, where
%
\begin{equation}
    \label{eqn:z}
    \Zone_{t,\tilde{t}^*} \equiv \frac{1}{h}\sum_{n=1}^{h}Y^*_{t,h(\tilde{t}^*-1)+n}, \quad \Ztwo_{t,\tilde{t}^*} \equiv \sum_{k=1}^{\tilde{\omega}}\Big|\Big|DFT\{Y^*_{t,h(\tilde{t}^*-1)+1},\ldots, Y^*_{t,h\tilde{t}^*}\}(k)\Big|\Big|^2.
\end{equation}
%
In words, $\Zone_{t,\tilde{t}^*}$ is the average acceleration vector over window $\tilde{t}^*$ and $\Ztwo_{t,\tilde{t}^*}$ is the squared two-norm of the component of the window that can be attributed to frequencies between one and $\tilde{\omega} \leq h-1$ periods per window (by the Plancherel theorem). More intuitively, $\Ztwo_{t,\tilde{t}^*}$ corresponds to the ``wiggliness" of the acceleration data for dive $t$ within window $\tilde{t}^*$. The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with the help of a subject-matter expert. Other DFT-based choices for window summary statistics are possible, including the dominant frequency and amplitude. Setting the stride length equal to the window size $h$ avoids the artificial residual correlation that overlapping windows would introduce and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting. By summarizing the data using a relevant transformation, this method reduces correlation in the raw observations due to any complex dependence structure.

The fine-scale HMM must be redefined since $\Z_t$ exists on a coarser scale than $Y^*_t$ itself. In particular, the length of $\Z_t$ is denoted as $\tilde{T}_t^*$ and there are only $\tilde{T}_t^*$ hidden states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$, denoted as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$. The fine-scale probability transition matrices  $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ are then applied directly to $\tilde{X}^*_t$ and $\Z_t$ rather than $X^*_t$ and $Y^*_t$. It then follows that the likelihood of this model is identical to that of the original HMM but with $Y^*$ replaced with $\Z$ and $X^*$ replaced with $\tilde{X^*}$. In order to clearly differentiate models, we refer to an HMM with $\Z$ as observations and $\tilde{X}$ as hidden states as an HMM-DFT.

### 2020-11-06 Report/Sections/simulation.tex
Fix Overleaf issues

%\section{Simulation Study}
We perform a simulation study based on data generated from the full CarHHMM-DFT as defined in Section \ref{subsec:model_selection} to evaluate each candidate model when the ground-truth is known. The parameters used to generate the data are loosely based on those estimated in the case study (see Table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity. In particular, we set $\Gamma$ such that the simulated data have equal numbers of short and long dives in expectation, the number of subdive states is set to $N^*=2$, and $\Zone_{t,t^*}$ is set to a scalar rather than a three dimensional vector. We then fit the data with four models: the full CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT. Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, empirical standard errors of parameter estimates, and fitting times. To assess the accuracy of uncertainty estimates, we also compare the empirical standard errors a given model's parameter estimates with the standard errors estimated using the inverse of the observed Fisher information under that model.

@@

We generate 500 independent data sets from the CarHHMM-DFT. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
%
Dive duration is Gamma distributed, and the coarse-scale emission parameters $\theta$ are $\mu^{(1)} = 20s$, $\sigma^{(1)} = 5s$, $\mu^{(2)} = 80s$, and $\sigma^{(2)} = 25s$. After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2T^*_t$ seconds of each dive are ignored. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left\{X^*_{t,1}, \ldots \right\}$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}$$ 
%
for dive types 1 and 2, respectively.
Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $Z^*_{t,t^*} = \left\{\Zone_{t,t^*}, \Ztwo_{t,t^*}\right\}$ as defined in Equation (\ref{eqn:z}). Recall that $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ are independent when conditioned on all dive types and subdive states. Each observation of acceleration $\Zone_{t,t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,t^*}$ is Gamma-distributed without explicit auto-correlation. In particular, we set the distributions of $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ with the following parameters:
%
\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.05s, \enspace \phi_A^{*(\cdot,1)} = 0.99, \\
    %
    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.1s, \enspace \phi_A^{*(\cdot,2)} = 0.95, \\
    %
    \mu_W^{*(\cdot,1)} = 10.10, \quad \sigma_W^{*(\cdot,1)} = 3.18, \\
    %
    \mu_W^{*(\cdot,2)} = 305.94, \quad \sigma_W^{*(\cdot,2)} = 17.49.
\end{gather*}
%
It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix. Figure \ref{fig:sim_data} shows one realization of $Z^*$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.

@@

We use several metrics to compare the candidate models, including its accuracy when decoding dive types and subdive states (see Table \ref{table:accuracy}). If $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define average dive decoding accuracy as the average value of $\Pr(X_{t} = x_{t}|Y,Y^*)$ across all simulated data sets and dives. Analogously, if $X^*_{t,t^*}$ is the unknown subdive state associated with a particular model and $x^*_{t,t^*}$ is the true simulated subdive state for dive $t$ and window $t^*$, then we define the average subdive decoding accuracy as $\Pr(X^*_{t,t^*} = x^*_{t,t^*}|Y,Y^*)$ averaged across all simulated data sets, dives, and windows. We find both the average dive decoding accuracy and the average subdive decoding accuracy using the forward-backward algorithm.

>>>>

%\section{Simulation Study}
We perform a simulation study based on data generated from the full CarHHMM-DFT as defined in Section \ref{subsec:model_selection} to evaluate each candidate model when the ground-truth is known. The parameters used to generate the data are loosely based on those estimated in the case study (see Table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity. In particular, we set $\Gamma$ such that the simulated data have equal numbers of short and long dives in expectation, the number of subdive states is set to $N^*=2$, and $\Zone_{t,\tilde t^*}$ is set to a scalar rather than a three dimensional vector. We then fit the data with four models: the full CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT. Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, empirical standard errors of parameter estimates, and fitting times. To assess the accuracy of uncertainty estimates, we also compare the empirical standard errors a given model's parameter estimates with the standard errors estimated using the inverse of the observed Fisher information under that model.

@@

We generate 500 independent data sets from the CarHHMM-DFT. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
%
Dive duration is Gamma distributed, and the coarse-scale emission parameters $\theta$ are $\mu^{(1)} = 20s$, $\sigma^{(1)} = 5s$, $\mu^{(2)} = 80s$, and $\sigma^{(2)} = 25s$. After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each dive are ignored. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \left\{X^*_{t,1}, \ldots \right\}$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}$$ 
%
for dive types 1 and 2, respectively.
Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z^*_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$ as defined in Equation (\ref{eqn:z}). Recall that $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are independent when conditioned on all dive types and subdive states. Each observation of acceleration $\Zone_{t,\tilde t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,\tilde t^*}$ is Gamma-distributed without explicit auto-correlation. In particular, we set the distributions of $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ with the following parameters:
%
\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.05s, \enspace \phi_A^{*(\cdot,1)} = 0.99, \\
    %
    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.1s, \enspace \phi_A^{*(\cdot,2)} = 0.95, \\
    %
    \mu_W^{*(\cdot,1)} = 10.10, \quad \sigma_W^{*(\cdot,1)} = 3.18, \\
    %
    \mu_W^{*(\cdot,2)} = 305.94, \quad \sigma_W^{*(\cdot,2)} = 17.49.
\end{gather*}
%
It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix. Figure \ref{fig:sim_data} shows one realization of $Z^*$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.

@@

We use several metrics to compare the candidate models, including its accuracy when decoding dive types and subdive states (see Table \ref{table:accuracy}). If $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define average dive decoding accuracy as the average value of $\Pr(X_{t} = \tilde x_{t}|Y,Y^*)$ across all simulated data sets and dives. Analogously, if $\tilde X^*_{t,\tilde t^*}$ is the unknown subdive state associated with a particular model and $\tilde x^*_{t,\tilde t^*}$ is the true simulated subdive state for dive $t$ and window $t^*$, then we define the average subdive decoding accuracy as $\Pr(\tilde X^*_{t,t^*} = \tilde x^*_{t,\tilde t^*}|Y,Y^*)$ averaged across all simulated data sets, dives, and windows. We find both the average dive decoding accuracy and the average subdive decoding accuracy using the forward-backward algorithm.
